---
title: "Make Linear River Grid"
author: "Peter Dudley"
date: "9/23/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```
# Make a Linear River Grid

This script takes 2 shape files, a point at the top of the river and a center line,  and makes a grid along the river.

Start with some inputs for the script. They are the default projection, the resolution, the max distance to make the grid out from the center line, and the 2 input shape files
```{r}
#____________________________________________________________________#
##### Inputs #####
# Set the desired EPGS code for the GIS files
EPSG = 32610
# Set the resolution you want
# This is how often a point is placed on the river line in m
# and how spaced the buffers are
resolution = 20
# Maximum lateral distance
# what is the max distance away from the center line 
# you want the lateral grid to extend in m
max_buffer = 500
# the paths and names of the input files
center_line = "../../../data/GIS/center_line_1/Center_Line.shp"
top_marker = "../../../data/GIS/top_point_1/Top_Point.shp"
#____________________________________________________________________#
```

Load the necessary libraries 
```{r message=FALSE, warning=FALSE}
# the simple features library for most of the shape file stuff
library(sf)
# leaflet for plotting shapefiles
library(leaflet)
# you know why
library(tidyverse)
# deal with most of the rater calculations
library(raster)
# to smooth the center line of the river
library(smoothr)
# the viridis color map
library(viridis)
# a faster way to do raster sampling with shape files
library(exactextractr)

# Make sure slects is the dplyr function
select = dplyr::select
```

Load Functions
This is a list of functions used to do the calculations as well as plot
```{r}
source("./scripts/Grid_Maker_Functions.R")
source("./scripts/Map_Maker_Functions.R")
```

Load the necessary data and do some pre processing of making a list of lateral distances for the grid
```{r}
shape_files = load_input_files(line = center_line ,
                             top = top_marker,
                             EPGS = EPSG)

distances_list = make_distances_list(resolution = resolution,
                                     buffer = max_buffer)
```

Display in the input files
```{r fig.align='center'}
make_leaflet_map(shape_files$line, 
                 type = "line")

make_leaflet_map(shape_files$top, 
                 type = "point")

```

First make a series of buffers which are the lateral grid dividers
```{r fig.align='center'}
buffers = make_buffers(distances = distances_list,
                       line = shape_files$line)

make_leaflet_map(buffers, 
                 type = "poly")
```

Make a shape used to tell left form right
```{r fig.align='center'}
large_buffer = make_large_buffer(distances = distances_list,
                                 line = shape_files$line)

make_leaflet_map(large_buffer, 
                 type = "poly")
```

Place sample points along the line
```{r fig.align='center'}
sample_points = make_sample_points(resolution = resolution,
                                   line = shape_files$line)

make_leaflet_map(sample_points, 
                 type = "point")
```

Make the Voronoi cells
```{r fig.align='center'}
vor_cells = make_vor_cells(points = sample_points,
                           top = shape_files$top,
                           resolution = resolution)

make_leaflet_map(vor_cells, 
                 type = "poly")
```

Combine the buffers and Voronoi cells to make the grid
```{r, warning=FALSE, fig.align='center'}
grid = make_grid(resolution = resolution,
                 cells = vor_cells,
                 buffers = buffers,
                 large_buffer = large_buffer)

make_leaflet_map(grid, 
                 type = "poly")
```


